---
title: "BUEC 333 - RMarkdown formatting instructions"
author: "Chris Muris and Tim Schulz"
date: "July 10, 2016"
output: html_document
---

#E4.1

To start, I load the data, using the [haven](https://github.com/hadley/haven/) package.

```{r}
library(haven)
growthdata = read_dta('Growth.dta')
```



##(a)
To plot the relationship between tradeshare and growth, I use the [ggplot](http://ggplot2.org/) package.

```{r}
require(ggplot2)
plot = ggplot(growthdata, aes(x=tradeshare, y=growth)) + geom_point()
plot
```

There appears to be a positive correlation between a country's trade share and its growth rate.

##(b)
Next, I Mark Malta's point in red, using the "colour" attribute in ggplot2, using code that I adapted from [cookbook-r](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#simple-color-assignment).

```{r}
plot + geom_point(aes(x=tradeshare, y=growth), growthdata[growthdata$country_name=='Malta',], colour='red')
```

Malta does indeed appear to be an outlier.

##(c)

Regress `growth` on `tradeshare`, using "lm".

```{r}
reg1 = lm(growth ~ tradeshare, growthdata)
```

The estimated intercept and slope coefficients are

```{r}
reg1$coefficients
```

For a country with `tradeshare=0.5`, we predict a growth rate of

```{r}
as.numeric(predict(reg1, data.frame(tradeshare=.5)))
```

For a country with `tradeshare=1`, we predict a growth rate of

```{r}
as.numeric(predict(reg1, data.frame(tradeshare=1)))
```

These predictions differ by 0.5 times the estimated regression coefficient on tradeshare.

##(d)

Next, we run the same regression, but now excluding Malta. In what follows, I am using code that was demonstrated during the computer lab of July 8, 2016.

```{r}
reg2 = lm(growth ~ tradeshare, growthdata[growthdata$country_name!='Malta',])
```

Without Malta, the estimated intercept and slope coefficients are

```{r}
reg2$coefficients
```

Without Malta, for a country with `tradeshare=0.5`, we predict a growth rate of

```{r}
as.numeric(predict(reg2, data.frame(tradeshare=.5)))
```

This prediction is close to the one we obtained when using the data frame with Malta.

Without Malta, for a country with `tradeshare=1`, we predict a lower growth rate of

```{r}
as.numeric(predict(reg2, data.frame(tradeshare=1)))
```

This prediction is roughly 0.3 below the one we obtained using Malta.